Adaptive evolution with aneuploidy and mutation

### 2022-03-02 ms/ms.tex
more fixes, priors sections still under construction

\paragraph{Prior distributions.} We used informative prior distributions for $w_{2n+1}=1-c+b$, $w_{2n+1^*}=(1+s)(1-c)+b$ and $w_{2n^*}=1+s$, which we estimated from growth curves data from mono-culture growth experiments previously reported by \citet[Figs. 3C, 4A, and S2]{Yona2012}.
We used \texttt{Curveball}, a method for predicting results of competition experiments from growth curve data~\citep[\href{https://curveball.yoavram.com}{curveball.yoavram.com}]{Ram2019}. Briefly, \texttt{Curveball} takes growth curves of two strains growing separately in mono-culture and predicts how they would grow in a mixed culture, that is, it predicts the results of a competition assay.
From these predictions, relative fitness values can be computed. Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix. We sampled 10,000 samples to use as a prior distribution~(\cref{fig:growth-curves}).
We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$ ~(\cref{fig:growth-curves}-A). In lieu of a better prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$. 
To increase computational efficiency, we also assumed $w(2n*)>w(2n+1*)>w(2n+1)>w(2n)$; running the inference without this assumption produced similar results.

Compared to other priors we tested, this prior produced lower WAIC, better posterior prediction plots, and more stable parameter estimates, as follows.
We tried to use additional growth curves. We used growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$. The same prior was used for $w_{2n*}/w_{2n+1*}$. This prior resulted in WAIC 0.32, compared to 0.27 with the above informative prior.
We also tried to use growth curves of \anwt\; and \euwt\; in \SI{30}{\celsius} ~(\cref{fig:growth-curves}-B) to estimate $1-c$ . This estimation assumes that the cost of aneuploidy is the same in \SI{39}{\celsius} and \SI{30}{\celsius}; this might be incorrect, but we only assumed this to generate a prior distribution for the fitness values. The prior for $b$ was taken the same as for $c$. This prior did not converge: no parameter sets were found with an approximate likelihood greater than zero.
% TODO IK it makes sense because bla bla (to explain)
Finally, we also tested a uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ and (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$. In both cases the algorithm failed to converge.

For the mutation rate, $\mu$, and aneuploidy rate, $\delta$, we used uninformative uniform priors, $\mu \sim \mathit{U}(10^{-9}\ ,\ 10^{-5})$ and $\delta \sim \mathit{U}(10^{-6} , 10^{-2})$. A wider mutation rate prior, $\mu \sim \mathit{U}(10^{-9}\ ,\ 10^{-5})$, produced similar results.
For a model without aneuploidy ($\delta=0$), we used a wider uniform prior for the mutation rate, $\mu \sim \mathit{U}(10^{-10},10^{-5})$.

@@

% Almost all done from here already

@@

\paragraph{Model checking and comparison.}
The single-locus model fits the data well: in simulations using the best-fit parameters (MAP estimate) \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\cref{fig:2n*-fixation}-B) and \anwt\ fixed in roughly 300 generations on average (green lines in \cref{fig:ppc-plot}), which agrees with experimental results.
% TODO IK replace "roughly 300" with the exact number.
Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\cref{fig:ppc-plot}). % TODO move somewhere else?
Furthermore, sensitivity analysis shows that changing a single parameter while keeping the rest fixed at the MAP estimate produces a worse fit to the data~(\cref{fig:sensitivity}).

@@

% TODO YR Read this
Next we tried the model with aneuploidy and a fixed mutation rate $\mu$. Hence $\mu$ is not well identifiable by the model (the inferred value has high variance with no strong correlation between it and other parameters), we tried the models where $\mu$=$10^{-5}$, $10^{-6}$ and $10^{-7}$. We saw that the inference posteriors were similar to the model with free $\mu$ parameter. However WAIC was better for the models with fixed $\mu$ ~(\autoref{table:WAIC}). Model with the $\mu$=$10^{-5}$ has the best WAIC, and better than the model without aneuploidy.

@@

@@

>>>>

\paragraph{Prior distributions.} We used informative prior distributions for $w_{2n+1}=1-c+b$, $w_{2n+1^*}=(1+s)(1-c)+b$ and $w_{2n^*}=1+s$, which we estimated from growth curves data from mono-culture growth experiments previously reported by \citet[Figs. 3C, 4A, and S2]{Yona2012}.
We used \texttt{Curveball}, a method for predicting results of competition experiments from growth curve data~\citep[\href{https://curveball.yoavram.com}{curveball.yoavram.com}]{Ram2019}. Briefly, \texttt{Curveball} takes growth curves of two strains growing separately in mono-culture and predicts how they would grow in a mixed culture, that is, it predicts the results of a competition assay.
From these predictions, relative fitness values can be computed. Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix. We sampled 10,000 samples to use as a prior distribution~(\cref{fig:growth-curves}).
We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$ ~(\cref{fig:growth-curves}-D). In lieu of a better prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$. 
To increase computational efficiency, we also assumed $w(2n*)>w(2n+1*)>w(2n+1)>w(2n)$; running the inference without this assumption produced similar results.

Compared to other priors we tested, this prior produced lower WAIC, better posterior prediction plots, and more stable parameter estimates, as follows. %TODO IK what do you mean by better posterior prediction and is more stable parameter estimates still rellevant?
We tried to use additional growth curves. We used growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$ ~(\cref{fig:growth-curves}-L). The same prior was used for $w_{2n*}/w_{2n+1*}$. This prior resulted in WAIC 71.91, compared to 67.87 with the above informative prior. % generated with fixed-mutation-rate-and-WAICs.ipynb TODO IK update the WAIC with 1000 reps run
We also tried to use growth curves of \anwt\; and \euwt\; in \SI{30}{\celsius} to estimate $1-c$ ~(\cref{fig:growth-curves}-H). This estimation assumes that the cost of aneuploidy is the same in \SI{39}{\celsius} and \SI{30}{\celsius}; this might be incorrect, but we only assumed this to generate a prior distribution for the fitness values. The prior for $b$ was taken the same as for $c$. This prior did not converge: no parameter sets were found with an approximate likelihood greater than zero. %TODO IK check that this model had no bugs
% TODO IK it makes sense because bla bla (to explain)
Finally, we also tested a uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ and (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$. In both cases the algorithm failed to converge.

For the mutation rate, $\mu$, and aneuploidy rate, $\delta$, we used uninformative uniform priors, $\mu \sim \mathit{U}(10^{-9}\ ,\ 10^{-5})$ and $\delta \sim \mathit{U}(10^{-6} , 10^{-2})$. A wider mutation rate prior, $\mu \sim \mathit{U}(10^{-9}\ ,\ 10^{-3})$, produced similar results.
% TODO IK, maybe wider mutation rate should be tried on a model without aneuploidy
%For a model without aneuploidy ($\delta=0$), we used a wider uniform prior for the mutation rate, $\mu \sim \mathit{U}(10^{-10},10^{-5})$.

@@

@@

\paragraph{Model checking and comparison.}
The single-locus model fits the data well: in simulations using the best-fit parameters (MAP estimate) \eumt\ fixed in 61\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\cref{fig:2n*-fixation}-B),
Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\cref{fig:ppc-plot}). % TODO move somewhere else?
Furthermore, sensitivity analysis shows that changing a single parameter while keeping the rest fixed at the MAP estimate produces a worse fit to the data~(\cref{fig:sensitivity}).

@@

@@

We also inferred model parameters under the assumption that the mutation rate in aneuploid cells, compared to euploid cells, is increased by a factor $\tau$=1, 33/32 (due to an additional chromosome), 2, 5, 10 or 100 (due to genetic instability).
We found that the posterior distribution was similar for $\tau=$1, 33/32, 2, and 5 (\cref{fig:tau}).
%% generated with MAP-and-hdi.ipynb
With $\tau=10$, the estimated mutation rate was slightly lower ($\mu=1.674\times10^{-6} (2.501\times10^{-7}-1.741\times10^{-6})$). 
With $\tau=100$, the estimated mutation rate was considerably lower ($\mu=3.81\times10^{-7} (1.508\times10^{-7}-4.995\times10^{-7})$) and the aneuploidy rate had higher variance ($\delta=7.824\times10^{-4} (6.616\times10^{-4}-1.462\times10^{-3})$). 
%TODO YR to reference to WAIC table also? By this table tau=100 is the best model
We cannot rule out an increased mutation rate in aneuploid cells, but we find no evidence that it does, and unless the effect is strong, it does not seem to affect our results.
We also checked the differences in fixation/loss times of  \euwt, \anwt and \eumt\  for the models with different $\tau$. We saw that the models can be distinguished using these parameters, especially $\tau=100$ ~(\cref{fig:tau-plots}).

% TODO YR Read this
Next we tried the model with fixed mutation rate $\mu$. Hence $\mu$ is not well identifiable by the model (the inferred value has high variance with no strong correlation between it and other parameters), we tried the models where mutation rate is fixed: $\mu$=$10^{-5}$, $10^{-6}$ and $10^{-7}$. We observed similar parameters inference comparing to the model with free $\mu$ parameter. However WAIC was better for the models with fixed $\mu$ ~(\autoref{table:WAIC}). The lowest WAIC got the model with $\mu$=$10^{-5}$ - 0.6, it is lower than \emph{no aneuploidy} model - 0.87, that implies higher predictive accuracy.

@@

% Fig tau plots
%% generated by diff-tau-plots.ipynb
\begin{figure}[h!]
  \centering
  \begin{subfigure}{0.75\textwidth}
      \includegraphics[width=\textwidth]{../figures/tau-a.pdf}      
  \end{subfigure}
  \begin{subfigure}{0.75\textwidth}
      \includegraphics[width=\textwidth]{../figures/tau-b.pdf}      
  \end{subfigure}
  \\
   \begin{subfigure}{0.75\textwidth}
      \includegraphics[width=\textwidth]{../figures/tau-c.pdf}      
  \end{subfigure}
   \caption{
    \textbf{Genotype fixations for single-locus models with different $\tau$.} For each model, $10,000$ simulations using MAP parameters were evaluated and a percentage of simulations with fixated genotypes  \euwt, \anwt\ and \eumt\ (plots A, B and C respectively) were measured in time. \anmt\ didn't rich fixation, therefore it doesn't appear in figures. We can see that $\tau=100$ can be distinguished using data of loss of \euwt\ (plot A) and fixation/loss of \anwt\ (plot B). It is harder to distinguish $1\leq\tau\leq10$, however the correlation exists.
  }
  \label{fig:tau-plots}
\end{figure}
